\name{}: machine learning over interaction networks and ontology-based background knowledge predicts drug--target interactions

### 2021-07-01 paper/main.tex
.

The most common scheme for drug--target interaction prediction is the
split over drug-target pairs \cite{Survey2018} where it may happen
that most drugs and targets that are including in the model's
validation and testing phase have also been included in the training
phase (as part of other drug--target pairs). This scheme is prone to a
number of biases. If the number of interactions for a drug or protein
are imbalanced, i.e., some drugs or proteins have many more
interactions than others, these will be seen more often during
training and they will likely also have more interactions in the
testing and validation sets; because some entities have more
interactions, i.e., they are more likely to interact, any model that
preferentially predicts these as interaction partners will improve its
predictive performance. While this accurately captures the
distribution, predicting based on biases in the number of interaction
partners is not desirable when applying the model to novel
entities. We have demonstrated that even the newly proposed ``na\"ive'' classifier that
makes predictions only by exploiting the imbalanced number of
interaction partners can achieve performance close to state of the art
methods (when measuring Macro AUC). When training a machine learning
model on such imbalanced data, it will eventually overfit to this
imbalance.  Splitting by entity (protein or drug) can reduce the
impact of this spurious correlation but not reduce it entirely,
because similar entities will still exhibit similar interaction
patterns. In our experiments, we observed the impact of splitting
training and evaluation sets by protein as a decrease in overall
performance (Macro AUC), providing some evidence that models trained
using this splitting scheme are less sensitive to overfitting to this
type of bias.

The way in which training and evaluation data is generated is related
to how the model is evaluated. An evaluation based on Macro AUC
evaluated the application scenario where a set of drugs and proteins
are given, and out of all possible pairs, the more likely interactions
need to be identified. However, this does not correspond to most
scenarios for drug repurposing where a drug that targets a {\em
  specific} protein (e.g., a protein involved in a disease) needs to
be identified. We use an evaluation measure based on micro-averages
per protein (Micro AUC$_p$) to evaluate this scenario, and we often
find substantial differences in predictive performance when evaluating
with Macro AUC and Micro AUC$_p$; generally, models that are trained
using a split over drug--target pairs perform worse in Micro AUC$_p$
than models that use a protein-based split, further providing evidence
that a drug--target split results in overfitting to dataset biases.

>>>>

The most common scheme for drug--target interaction prediction is the
split over drug-target pairs \cite{Survey2018} where it may happen
that most drugs and targets that are including in the model's
validation and testing phase have also been included in the training
phase (as part of other drug--target pairs). This scheme is prone to a
number of biases. If the number of interactions for a drug or protein
are imbalanced, i.e., some drugs or proteins have many more
interactions than others, these will be seen more often during
training and they will likely also have more interactions in the
testing and validation sets; because some entities have more
interactions, i.e., they are more likely to interact, any model that
preferentially predicts these as interaction partners will improve its
predictive performance. While this accurately captures the
distribution, predicting based on biases in the number of interaction
partners is not desirable when applying the model to novel
entities. We have demonstrated that even the newly proposed
``na\"ive'' classifier that makes predictions only by exploiting the
imbalanced number of interaction partners can achieve performance
close to state of the art methods (when measuring Macro AUC). When
training a machine learning model on such imbalanced data, it will
eventually overfit to this imbalance.  Splitting by entity (protein or
drug) can reduce the impact of these spurious correlations but not
reduce it entirely, because similar entities will still exhibit
similar interaction patterns. In our experiments, we observed the
impact of splitting training and evaluation sets by protein as a
decrease in overall performance (Macro AUC), providing some evidence
that models trained using this splitting scheme are less sensitive to
overfitting to this type of bias.

The way in which training and evaluation data is generated is related
to how the model is evaluated. An evaluation based on Macro AUC
evaluated the application scenario where a set of drugs and proteins
are given, and out of all possible pairs, the more likely interactions
need to be identified. However, this does not correspond to most
scenarios for drug repurposing where a drug that targets a {\em
  specific} protein (e.g., a protein involved in a disease) needs to
be identified. We introduce an evaluation measure based on
micro-averages per protein (Micro AUC$_p$) to evaluate this scenario,
and we often find substantial differences in predictive performance
when evaluating with Macro AUC and Micro AUC$_p$; generally, models
that are trained using a split over drug--target pairs perform worse
in Micro AUC$_p$ than models that use a protein-based split, further
providing evidence that a drug--target split results in overfitting to
dataset biases.
